---
title: "SC398 Homework 10"
author: Frank Driscoll, Cam Garfield, Cole Guerin, Hannah Hartnett
date: "Due Thursday, May 6, 2021 on Moodle"
output: 
  html_document:
    toc:  true
    toc_float:  true
    code_folding:  show
runtime: shiny
---

#  Homework 10: Interaction with Shiny and Plotly


***
***

# Problem 0

## Project dataset and office hour
Link to dataset: 

https://acleddata.com/data-export-tool/

https://raw.githubusercontent.com/hannahhartnett/sc398/main/US-ACLED.csv?token=APJP4N4Q2STICLA7OPLGY3DASLSQ4

Dataset Description: 

  The ACLED dataset provides information about political violence and protests in various regions of the world; we have chosen the United States in the year 2020. The data are collected weekly from secondary sources such as news reports, and independently coded (using the same coding rubric) by three researchers. It is important to note that each record represents a different date, location, agent, and event, ie. if one protest lasts for three days in a specific location, it will be recorded as three different units in the dataset.  
  There are 23238 records in our dataset with 31 variables. The key variables include event date, type of event (peaceful protest, protest with intervention, violence against civilians, etc.), actors involved in the event, location of the event, news source used to code the event, number of fatalities, and notes taken about the event by the researchers. 

Questions for dataset: 

1. Did peaceful protests increase or decrease in the United States after the beginning of lockdown orders in March 2020?

2. Was there an increase in protests 'with intervention' in the United States after the murders of George Floyd, Ahmaud Arbery, Breonna Taylor, and other Black Americans in the spring of 2020? Where did these protests occur?

3. How did the type and location of protests change over the course of the presidential election? Can we identify pro-Trump and pro-Biden protests?


Signed up for office hours on: 

Wednesday 5/12 at 1:45pm


## Shiny tutorial

Answer: We've all read/worked through the tutorial.

```{r}
faithful$week <- rep(c("Week 1", "Week 2"), each = 136)
```

# Problem 1: Shiny Outputs

a.

```{r base_r, echo=FALSE}
shinyApp(
  
  ui = fluidPage(
    inputPanel(
      selectInput("n_breaks", label = "Number of bins:",
                  choices = c(10, 20, 35, 50), selected = 20),
  
      sliderInput("bw_adjust", label = "Bandwidth adjustment:",
                  min = 0.2, max = 2, value = 1, step = 0.2)
    ),
    
    plotOutput("geyser_plot")
  ),
  
  server = function(input, output) {
    output$geyser_plot <- renderPlot({
      hist(faithful$eruptions, probability = TRUE,
           breaks = as.numeric(input$n_breaks),
           xlab = "Time (minutes)", ylab = "Density",
           main = "Geyser eruption duration")
      
      dens <- density(faithful$eruptions, adjust = input$bw_adjust)
      lines(dens, col = "blue")
    })
  },
  
  options = list(height = 550)
)
```

b.

```{r ggplot, echo=FALSE, warning=FALSE, message=FALSE}
library(ggplot2)
shinyApp(
  
  ui = fluidPage(
    inputPanel(
      selectInput("n_breaks", label = "Number of bins:",
                  choices = c(10, 20, 35, 50), selected = 20),
  
      sliderInput("bw_adjust", label = "Bandwidth adjustment:",
                  min = 0.2, max = 2, value = 1, step = 0.2)
    ),
    
    plotOutput("geyser_plot")
  ),
  
  server = function(input, output) {
    output$geyser_plot <- renderPlot({
      ggplot(data = faithful, mapping =  aes(x = eruptions)) +
        geom_histogram(mapping = aes(y = ..density..), bins = as.numeric(input$n_breaks), fill = I("grey"), color = I("black")) +
        geom_density(adjust = input$bw_adjust, color = "blue") +
        labs(x = "Time (minutes)", y = "Density",
           title = "Geyser eruption duration",
           caption = "Source: W. Härdle") +
        theme_classic() +
        scale_x_continuous(breaks=c(1.5,2,2.5,3,3.5,4,4.5,5)) +
        scale_y_continuous(breaks=c(0,0.1,0.2,0.3,0.4,0.5,0.6,0.7))
    })
  },
  
  options = list(height = 550)
)
```

***
***

# Problem 2: Shiny Inputs

```{r more_inputs, echo=FALSE, warning=FALSE, message=FALSE}
shinyApp(
  
  ui = fluidPage(
    inputPanel(
      selectInput("n_breaks", label = "Number of bins:",
                  choices = c(10, 20, 35, 50), selected = 20),
  
      sliderInput("bw_adjust", label = "Bandwidth adjustment:",
                  min = 0.2, max = 2, value = 1, step = 0.2),
      
      checkboxInput("do_rug", label = "Rug Plot"),
  
      radioButtons("which_variable", label = "Which variable?",
                   choices = c("Eruptions" = "eruptions",
                               "Waiting" = "waiting"))
    ),
    
    plotOutput("geyser_plot")
  ),
  
  server = function(input, output) {
    output$geyser_plot <- renderPlot({
      p2 <- ggplot(data = faithful, mapping =  aes_string(x = input$which_variable)) +
        geom_histogram(mapping = aes(y = ..density..), bins = as.numeric(input$n_breaks), fill = I("grey"), color = I("black")) +
        geom_density(adjust = input$bw_adjust, color = "blue") +
        labs(x = "Time (minutes)", y = "Density",
           title = paste("Geyser", input$which_variable, "duration"),
           caption = "Source: W. Härdle") +
        theme_classic()
      
      if(input$do_rug) {
        p2 <- p2 + geom_rug()
      }
      
      p2
      
    })
  },
  
  options = list(height = 650)
)
```

***
***

# Problem 3: Plotly

*What information does Plotly display when you hover the mouse over a point? How about when you hover over the density contours but not over a point?*

Answer: When you hover over a point, Plotly displays the eruption and waiting time corresponding to the point. When you hover over the density contours, Plotly displays the contour level.

*What happens to the other facet? How about if you click & drag only horizontally*

Answer: When you click and drag vertically, Plotly has the plots for both facets zoom in, but when you click and drag horizontally, Plotly only has the plot for the facet you clicked on zoom in.

```{r plotly, echo=FALSE, warning=FALSE, message=FALSE}
library(plotly)

p3 <- ggplot(faithful, aes(x = eruptions, y = waiting)) +
  stat_density_2d(aes(fill = ..level..), geom = "polygon") + 
  xlim(1, 6) + ylim(40, 100) +
  geom_point() +
  facet_grid(~week)

ggplotly(p3)
```

***
***

# Problem 4: Plotly within Shiny

```{r plotly_in_shiny, echo=FALSE, warning=FALSE, message=FALSE}
## COPY YOUR SOLUTION TO PROBLEM 2 HERE, THEN MAKE THE REQUESTED CHANGES ##
shinyApp(
  
  ui = fluidPage(
    inputPanel(
      selectInput("n_breaks", label = "Number of bins:",
                  choices = c(10, 20, 35, 50), selected = 20),
  
      sliderInput("bw_adjust", label = "Bandwidth adjustment:",
                  min = 0.2, max = 2, value = 1, step = 0.2),
      
      checkboxInput("do_rug", label = "Rug Plot"),
  
      radioButtons("which_variable", label = "Which variable?",
                   choices = c("Eruptions" = "eruptions",
                               "Waiting" = "waiting"))
    ),
    
    plotlyOutput("geyser_plot")
  ),
  
  server = function(input, output) {
    output$geyser_plot <- renderPlotly({
      p4 <- ggplot(data = faithful, mapping =  aes_string(x = input$which_variable)) +
        geom_histogram(mapping = aes(y = ..density..), bins = as.numeric(input$n_breaks), fill = I("grey"), color = I("black")) +
        geom_density(adjust = input$bw_adjust, color = "blue") +
        labs(x = "Time (minutes)", y = "Density",
           title = paste("Geyser", input$which_variable, "duration"),
           caption = "Source: W. Härdle") +
        theme_classic()
      
      if(input$do_rug) {
        p4 <- p4 + geom_rug()
      }
      
      p4
      
    })
  },
  
  options = list(height = 650)
)
```

***
***
