---
title: "United States Protest Events in 2020"
output: 
  flexdashboard::flex_dashboard:
    orientation: columns
    vertical_layout: scroll
    theme: journal
runtime: shiny
---

<style type="text/css">

.chart-title {  /* chart_title  */
   font-size: 30px;
   font-family: News Cycle;

</style>

```{r}
library(flexdashboard)
library(shiny)
library(readr)
library(dplyr)
library(ggplot2)
library(leaflet)
library(plotly)
library(shinythemes)
library(sysfonts)
library(showtext)

acled <- read_csv('acled.csv')
death_count <- read_csv('death_count.csv')


font_add_google('News Cycle', 'news_cycle')
showtext_auto()

plot_theme <-  theme_gray() + # White background, black and white theme
  theme(axis.text = element_text(size = 10, color = "azure4"),
        text = element_text(size = 10, face = "bold", color = "azure4", family = 'news_cycle'))
```




Event Type and Location
================================

Column {data-width=350}
-----------------------------------------------------------------------

In the year of the 'new normal,' we have experienced many momentous and novel events: a global pandemic; increased awareness of national racial injustices, specifically on the part of the police force, as a result of the murders of countless Black Americans; and a contentious presidential election. This interactive app is intended as an accompaniment to year-end news summaries, to aid investigation of protest events as a response to these significant new phenomena. How have Americans gathered to react to national events in 2020? The app was created using data from the Armed Conflict Location and Event Data Project: https://acleddata.com/curated-data-files/. This data contains information on both armed conflicts and protest events, so peaceful protests are also included.

### Events by State

```{r}
library(lubridate)
acled_time <- acled %>% group_by(admin1, event_date) %>% summarize(count = n())
# view by state
library(shiny)
ui <- shinyUI(
  fluidPage(theme = shinytheme("cerulean"),
    sidebarLayout(
      sidebarPanel(
        uiOutput("stateInput")
      ),
      mainPanel(
        plotOutput("statePlot", 
                   height = '250px')
      )
    )
  ))
server <- shinyServer(function(input, output) {
  output$stateInput <- renderUI({
    selectInput(
      inputId = "state",
      label = "Select a State:",
      choices = levels(as.factor(acled_time$admin1))
    )
  })
  output$statePlot <- renderPlot({
    ggplot(acled_time %>% filter(admin1 == input$state), aes(x = event_date, y = count)) + geom_line() +
      labs(x = 'Event Date',
           y = 'Count', 
           title = 'Total Number of Protest Events per Day, by State') +plot_theme
  })
})
shinyApp(ui = ui, server = server, options = list(height = 400))
```


#### Key Trends
- The number of protests per day peaks earlier in Minnesota and Georgia, the location of the murders of George Floyd and Ahmaud Arbery, respectively, than most other states. 
- The overall number of protests in Washington D.C. is fewer than in most other states, but it is greatest in the fall, at the peak of the election season for the 2020 Presidential Election.


Column {data-width=650, style="height:200pc;"}
-----------------------------------------------------------------------

### Event Locations

```{r}
shinyApp(

  ui = fluidPage(theme = shinytheme("cerulean"),
                 titlePanel("Protest Events During 2020"),
                 sidebarLayout(
                   sidebarPanel(
                   selectInput("event", h3("Select an Event Type:"), 
                       choices = levels(as.factor(acled$event_type)), selected = "Protests"),
                   sliderInput("slider", "Dates:",
                               min = as.Date("2020-01-01","%Y-%m-%d"),
                               max = as.Date("2020-12-31","%Y-%m-%d"),
                               value = c(as.Date("2020-04-01"), as.Date("2020-04-30")),
                               timeFormat = "%Y-%m-%d")

                 ),
                 mainPanel(plotlyOutput("sub_eventPlot",
                                        height = '292px'),
                 leafletOutput("event_map")))
  ),

  server = function(input, output) {
    output$event_map <- renderLeaflet({
      
      filtered_data <- acled %>% filter(event_date <= input$slider[2] & event_date >= input$slider[1] & event_type == input$event)
      
      if (nrow(filtered_data != 0)) {
        leaflet(acled %>% filter(event_date <= input$slider[2] & event_date >= input$slider[1] & event_type == input$event)) %>%
          setView(lng = -114.948, lat = 44.212, zoom = 3) %>%
          addTiles() %>%
          addCircles(lng = ~longitude, lat = ~latitude, label = ~lapply(lab, htmltools::HTML), popup = ~notes)
      } else {
        leaflet(acled %>% filter(event_date <= input$slider[2] & event_date >= input$slider[1] & event_type == input$event)) %>%
          setView(lng = -114.948, lat = 44.212, zoom = 3) %>%
          addTiles()
      }
      })

      output$sub_eventPlot <- renderPlotly({
        ggplotly(ggplot(acled %>% filter(event_type == input$event & (event_date <= input$slider[2] & event_date >= input$slider[1])), aes(x = sub_event_type)) + geom_bar() + coord_flip() +
                   scale_fill_manual(values =c('blue')) +
                   labs(x = 'Sub Event Type',
                        y = 'Count',
                        title = 'Count of Sub-Event Type'),
                 tooltip = 'y')})
    },

    options = list(height = 775)
    )
```

#### Key Trends
- There were a wide variety of different protest events in the US in 2020; the vast majority were peaceful protests, but we also see a large number of protests with intervention and violent demonstrations.

#### How to Use the Map

This map displays all protest events that occurred in the United States over the course of 2020. Each point represents one day on which the event occurred. Hovering over a point shows the date, type of event, and location, and clicking on a point displays notes about what took place. The slider changes the time frame that the plot displays, and choosing the Sub-Event Type to the left will also filter this map accordingly.


Event Actors
================================


Column {data-width=650, style="height:1000pc;"}
-----------------------------------------------------------------------

### Actor Overall Event Chart
```{r}
actorList <- unique(acled$assoc_actor_label1)
ui <- shinyUI(
  fluidPage(
    titlePanel("Actor Groups"),
    sidebarLayout(
      sidebarPanel(
         checkboxGroupInput("actor", "Choose Actors", choices = actorList, selected = c("AltRight", "Democrats", "Leftist", "Progressive", "Republicans"))
      ),
      mainPanel(
        plotOutput("actorPlot", 
                   height = '450px')
      )
    )
  ))
server <- shinyServer(function(input, output) {
  # output$actorInput <- renderUI({
  #   selectInput(
  #     inputId = "actor",
  #     label = "Select an Actor:",
  #     choices = levels(as.factor(acled$assoc_actor_label1))
  #   )
  # })
  output$actorPlot <- renderPlot({
    ggplot(acled %>% filter(assoc_actor_label1 == c(input$actor)), aes(x = assoc_actor_label1, fill = assoc_actor_label1)) + geom_bar() +
      theme(legend.title = element_text(size = 14),
  legend.text = element_text(size = 10)) +
      labs(x = 'Actor Groups',
           y = 'Count', 
           fill = 'Actor Groups',
           title = 'Total Number of Events Participated in by Actor Groups') + plot_theme
  })
})
shinyApp(ui = ui, server = server, options = list(height = 750))
```

### Actor and Sub Event Type Charts
```{r}


ui <- shinyUI(
  fluidPage(theme = shinytheme("cerulean"),
    titlePanel("Actors and Event Types"),
    sidebarLayout(
      sidebarPanel(
        uiOutput("actorInput2")
      ),
      mainPanel(
        plotlyOutput("actorPlot2", 
                   height = '350px')
      )
    )
  ))

server <- shinyServer(function(input, output) {

  output$actorInput2 <- renderUI({
    selectInput(
      inputId = "actor2",
      label = "Select Actors:",
      choices = levels(as.factor(acled$assoc_actor_label1))
    )
  })

  output$actorPlot2 <- renderPlotly({
    p3 <- ggplot(acled %>% filter(assoc_actor_label1 == input$actor2), 
      aes(x = sub_event_type))+
      plot_theme+
      geom_bar() +
      labs(x = 'Sub Event Type',
           y = 'Count', 
           title = 'Count of Sub-Event Types by Actors') +
      theme(axis.text.x = element_text(angle = 35, hjust = 1, vjust = 0.5))
      
    ggplotly(p3, tooltip = "y")
    
  })
  })


shinyApp(ui = ui, server = server, options = list(height = 425))
```

#### Key Trends
  - Black Lives Matter and other groups protesting for the end of racist institutions, policies, etc. were by far the most active group in 2020
  - Many other groups were also active; this shows great potential for a study to be done comparing actor activity over time
  
#### Information About Data
  Actor groups are categorized by affliation. So, BLM protestors, African American designated groups, the ACLU, and NAACP are all incorporated under the tag "BLM-Related". Migrant Rights includes ABOLISH ICE protestors, DREAMers, Refugees, La Resistancia, and other migrant identified groups. Similarly, politically affiliated groups included Republican identified groups and Democrat identified groups. We also categorized a few other politically based groups such as Progressive organizations like the MoveOn movement, the Indivisible Movement, the Sunrise Movement and Democratic Socialists for America as well as more leftist operating groups such as Communists, Socialists, ANTIFA and other anti-fascist identified groups. Similarly, groups such as the Boogaloo Boys, Proud Boys, Tea Party, White Nationalists, QAnon, and Patriot Party groups were all lumped together under the tag AltRight. The remaining tags were all generated by the keywords implied in their names, with those that remained grouped together as "Other".



Fatalities
================================


Column {data-width=650, style="height:1000pc;"}
-----------------------------------------------------------------------

### Fatalities by Actor

```{r}
tab1 <- select(acled, groups, fatality)
death_table <- tab1 %>% 
  group_by(groups) %>%
  count(fatality)
plotType <- function(input) {
  switch(input,
         Count = "stack",
         Proportional = "fill")
}
ui <- shinyUI(
  fluidPage(theme = shinytheme("cerulean"),
    titlePanel("Group"),
    sidebarLayout(
      sidebarPanel(
        sliderInput("min_n", label = "Minimum Number of Incidents by Group",
                  min = 0, max = 20, value = 20, step = 2),
        radioButtons("pType", "Choose plot type:",
                 list("Count", "Proportional"))
      ),
      mainPanel(
        plotOutput("incident_plot")
      )
    )
  ))
  
  
server <- shinyServer(function(input, output) {
  output$incident_plot <- renderPlot({
    colorset = c('Other'='#e6194B','Police'='#ffe119','Police and Other'='#469990',
                 'Police and Sole Perpetrator'='#4363d8','Protesters'='#800000',
                 'Protesters and Other'='#911eb4','Protesters and Police'='#42d4f4',
                 'Protesters and Rioters'='#bfef45','Protesters and Sole Perpetrator'='#fabed4',
                 'Rioters'='#3cb44b','Rioters and Other'='#dcbeff','Rioters and Police'='#aaffc3',
                 'Rioters and Sole Perpetrator'='#f58231','Sole Perpetrator'='#808000',
                 'Sole Perpetrator and Other'='#000075')
    ggplot(death_table %>% filter(n>input$min_n), aes(x=fatality, y=n, fill=groups))+
      geom_bar(position = plotType(input$pType), stat = "identity")+
      scale_fill_manual(values=colorset)+
      labs(x = "Fatality?", y = "Number of Incidents", fill = "Groups Involved") + plot_theme
  })
})
shinyApp(ui, server, options = list(height = 500))
```

#### Key Trends
- Incidents with no fatalities vastly outnumber incidents with fatalities, as can be seen by the count figure.
- The vast majority of non-fatality incidents involve only protestors, while the vast majority of incidents with a fatality involved Police officers.

Column {data-width=400, style="height:35pc;"}
-----------------------------------------------------------------------

### Total Fatalities


```{r}
shinyApp(
  
  ui = fluidPage(theme = shinytheme("cerulean"),
    plotlyOutput("page2chartB")
  ),
  
  
  
  server = function(input, output) {
    
    output$page2chartB <- renderPlotly({
deaths_by_month <- 
  ggplot(death_count, aes(x = yearmon, y = n)) + geom_col() + plot_theme+
  labs(x = "Month", y = "Number of Deaths", title = "Number of Total Deaths by Month") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 0.5))
deaths_by_month <- deaths_by_month + aes(text = paste0("</br> Number of Incidents with Fatalities: ", n,
                                                      "</br> Percent of Total Incidents: ", 
                                                      paste(round(PctTot,2),"%")))
ggplotly(deaths_by_month, tooltip = "text")
    })
  }
)
```

#### Key Trends
- Incidents involving a fatality peaked in the summer of 2020 with June being the month with the most fatalities.
- Incidents involving a fatality make up a small proportion of total incidents by month.

Black Lives Matter Protests
================================


Column {data-width=650, style="height:200pc;"}
-----------------------------------------------------------------------

### Black Lives Matter Protests

```{r}
shinyApp(
  
  ui = fluidPage(theme = shinytheme("cerulean"),
    titlePanel("Growth of Black Lives Matter Movement"),
    sidebarLayout(
      sidebarPanel(
      radioButtons("time", label = "Time Frame:",
                   choices = c("Before George Floyd's Death" = "Before George Floyd's Death",
                               "Week After George Floyd's Death" = "Week After George Floyd's Death",
                               "June" = "June",
                               "July" = "July",
                               "August" = "August",
                               "September" = "September",
                               "October" = "October",
                               "November" = "November",
                               "December" = "December"))
      
    ),
    mainPanel(leafletOutput("event_map"))
  )),
  
  
  
  server = function(input, output) {
    
    output$event_map <- renderLeaflet({

if (input$time == "Before George Floyd's Death") {
  min_date = "2020-01-01"
  max_date = "2020-05-24"
} else if (input$time == "Week After George Floyd's Death") {
  min_date = "2020-05-25"
  max_date = "2020-05-31"
} else if (input$time == "June") {
  min_date = "2020-06-01"
  max_date = "2020-06-30"
} else if (input$time == "July") {
  min_date = "2020-07-01"
  max_date = "2020-07-31"
} else if (input$time == "August") {
  min_date = "2020-08-01"
  max_date = "2020-08-31"
} else if (input$time == "September") {
  min_date = "2020-09-01"
  max_date = "2020-09-30"
} else if (input$time == "October") {
  min_date = "2020-10-01"
  max_date = "2020-10-31"
} else if (input$time == "November") {
  min_date = "2020-11-01"
  max_date = "2020-11-30"
} else if (input$time == "December") {
  min_date = "2020-12-01"
  max_date = "2020-12-31"
}
      
      leaflet(acled %>% filter(grepl("BLM", assoc_actor_1)) %>% filter(event_date <= max_date & event_date >= min_date)) %>%
        addTiles() %>%
        setView(lng = -114.948, lat = 44.212, zoom = 3) %>%
        addCircles(lng = ~longitude, lat = ~latitude, label = ~lapply(lab, htmltools::HTML), popup = ~notes)
      
      
    })
  },
  
  options = list(height = 500)
)
```

#### Key Trends
- The Black Lives Matter Movement has grown tremendously in the United States over the course of 2020.
- Based on the time frames, changes in frequency of protests associated with BLM are easily visible.
- After George Floyd's death, the number of BLM protests sees a dramatic spike and this continues into June.
- As time wears on for the rest of the year, the amount of protests gradually drops, but the protest events are still much more frequent than in the first five months of 2020.
- The protests associated with George Floyd's death raised awareness and allowed more people to join the cause and protest against other instances of violence against black lives.
- In the later months of 2020 not all points are directly reference George Floyd in the notes, and instead focus on other American citizens and their stories.


