---
title: |
    | How Has Major League Baseball Changed Over Time?
    | Exploring the Lahman's Baseball Database
author:
  - name: Frank Driscoll
  - name: Cam Garfield
  - name: Cole Guerin
  - name: Hannah Hartnett
    
  
affiliation:
  - address: Statistical Graphics, Colby College
column_numbers: 4
logoright_name: https&#58;//raw.githubusercontent.com/brentthorne/posterdown/master/images/betterhexlogo.png
logoleft_name: https&#58;//raw.githubusercontent.com/brentthorne/posterdown/master/images/betterhexlogo.png
output: 
  posterdown::posterdown_html:
    self_contained: false
bibliography: packages.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(Lahman)
library(dplyr)
library(reshape2)
library(ggplot2)
library(gridExtra)
library(viridis)

data(People)
people <- People
head(People)

batting <- Batting
head(batting)

pitching <- Pitching
head(pitching)

fielding <- Fielding
head(fielding)

sums <- batting %>% 
  group_by(playerID) %>% 
  summarise(ab_sum = sum(AB), k_sum = sum(SO), hr_sum = sum(HR))
sums$hr_ab = sums$hr_sum/sums$ab_sum
sums$k_ab = sums$k_sum/sums$ab_sum

by_ID <- sums %>% 
  filter(ab_sum > 1000) %>%
  filter(k_ab>0) %>% 
  filter(hr_ab>0)
  
by_year <- batting %>%
  group_by(yearID) %>%
  summarise(ab = sum(AB), k = sum(SO), hr = sum(HR))

by_year$hr_ab = by_year$hr/by_year$ab
by_year$k_ab = by_year$k/by_year$ab

# establish theme of graphs
baseball_theme <-  theme_gray() + # White background, black and white theme
  theme(axis.text = element_text(size = 10, color = "azure4"),
        text = element_text(size = 10, face = "bold", color = "azure4", family = 'Georgia'))
```

# Introduction



# The Database

Lahman's Baseball Database is part of the Comprehensive R  Archive Network. The updated version of the database contains complete batting and pitching statistics from 1871 to 2020, plus fielding statistics, standings, team stats, managerial records, post-season data, and more.

# Looking at Continuous Hit Data

```{r, warning = FALSE, message  = FALSE, echo = FALSE, fig.height = 6, out.width='100%', fig.align = "center"}

# Hannah plot 1
# take continuous variables
batting_continuous <- Batting %>% select(G, AB, R, H, X2B, X3B, HR, RBI, SB, CS, BB, SO, IBB, HBP, SH, SF, GIDP)

cormat <- round(cor(batting_continuous, use = 'complete.obs'), 2)


library(reshape2)
melted_cormat <- melt(cormat)


reorder_cormat <- function(cormat){
  # Use correlation between variables as basis for distance
  dd <- as.dist(1-abs(cormat))
  hc <- hclust(dd)
  cormat <- cormat[hc$order, hc$order]
  return(cormat)
}

# Function to get lower triangle of corr matrix
get_lower_tri <- function(cormat){
  cormat[upper.tri(cormat)] <- NA
  return(cormat)
}

lo_tri_reordered <- cormat %>%
  reorder_cormat %>%
  get_lower_tri %>%
  melt


ggplot(data = lo_tri_reordered, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile() +
  labs(x = 'Batting Variable',
       y = 'Batting Variable',
       title = 'Correlation Heat Map for Continuous Batting Variables',
       fill = 'Correlation') +
  baseball_theme +
  theme(axis.text.x = element_text(angle = 45)) +
  scale_fill_gradient2(
  low = "indianred1",
  mid = "white",
  high = "navy",
  midpoint = 0,
  limits = c(-1, 1))

```

# Exploring Hit Type

```{r, warning = FALSE, message  = FALSE, echo = FALSE, fig.height = 4, out.width='100%', fig.align = "center"}
# Cole Plot 1

batting <- Batting %>%
  # select the variables that we want left after we filter the data
  select(yearID, H, X2B, X3B, HR) %>%
  # select the years from 1871+
  filter(yearID >= 1871) %>%
  group_by(yearID) %>%
  #  summarise_each(funs(sum(., na.rm=TRUE))) %>%
  summarise_all(funs(sum(., na.rm=TRUE))) %>%
  # we summarize by year, and then na.rm takes care of 0's in the data
  mutate(X1 = H-(X2B+X3B+HR)) %>% #create a column for singles
  # we eventually want these as a percentage of hits, so we can do the math now
  mutate(Single = X1/H*100) %>%
  mutate(Double = X2B/H*100) %>%
  mutate(Triple = X3B/H*100) %>%
  mutate(HomeRun = HR/H*100)

bat <- batting %>%
  select(yearID, Single, Double, Triple, HomeRun)

bat_long <- melt(bat, id.vars = c("yearID"))

hitsperyear <- ggplot(bat_long, aes(x=yearID, y= value, col=variable)) +
  geom_line() +
  xlab("Major League Baseball Season") +
  ylab("Percentage") +
  ggtitle("Hits by Type in Major League Baseball") +
  scale_x_continuous(breaks = c(1870, 1885, 1900, 1915, 1930, 1945,
                                1960, 1975, 1990, 2005, 2020 )) +
  scale_y_continuous(breaks = c(0, 25, 50, 75, 100))+
  guides(colour=guide_legend(reverse=TRUE,
                             aes(ggtitle= "Type of Hit"))) +
  baseball_theme +
  scale_colour_manual(values = c('navy', 'indianred1', 'palegreen4', 'mediumpurple3'))


hitsperyear
```

It looks like that baseball has adapted over time so that less singles are being hit, while more doubles and home runs are being hit, and triples are becoming  less common. There is a particularly large increase in home runs in the last 10 years, and a particularly large decrease in singles in the last ten years.

# Batting Average getting Lower?

```{r, warning = FALSE, message  = FALSE, echo = FALSE, fig.height = 6, out.width='100%', fig.align = "center"}
# Cole Plot 2

S <- summarize(group_by(Batting, yearID, playerID),
               H=sum(H), AB=sum(AB),
               BB=sum(BB), HBP=sum(HBP, na.rm=TRUE),
               SF=sum(SF, na.rm=TRUE))

ST <- summarize(group_by(Teams, yearID),
                Games = round(mean(W + L)),
                min.AB = 3 * Games)
S2 <- merge(S, ST, by="yearID")
S.regular <- filter(S2, AB >= min.AB)

SD.data <- summarize(group_by(S.regular, yearID),
                     SD.AVG = sd(H / AB),
                     SD.OBP = sd((H + BB + HBP) / (AB + BB + HBP + SF)))

sd_avg <-  ggplot(SD.data, aes(yearID, SD.AVG)) + geom_point() +
  geom_smooth(colour = 'navy', se=FALSE, span=0.35) +
  labs(title = "Standard Deviations of AVG of Regular Hitters",
       x = "Year", y = "SD of Batting Average") +
  baseball_theme

sd_obp <- ggplot(SD.data, aes(yearID, SD.OBP)) +
  geom_point() + geom_smooth(colour = 'navy', se=FALSE, span=0.35) +
  labs(title = "Standard Deviations of OBP of Regular Hitters",
       x = "Year", y = "SD of On Base Percentage") +
  baseball_theme
grid.arrange(sd_avg, sd_obp, ncol = 1)
```

There was a recent study about the disappearance of the 0.400 batting average in baseball. It was theorized that this is because the batting average among “regular” players is steadily decreasing overtime, and that great hitters are moving closer to the average because of this smaller variation, so less baseball athletes have been able to reach the 0.400 batting average mark. From the top plot, we can see that the standard deviation of batting average for so called “regular” players has decreased over time, but it has been very consistent in the last 60 years. A different batting statistic, that has seen more of a drop recently, is the standard deviation for on base percentage which is shown in bottom plot.

# Sources

Lahman's Baseball Database 

http://www.seanlahman.com/baseball-archive/statistics/
