---
title: |
    | How Has Major League Baseball Changed Over Time?
    | Exploring the Lahman's Baseball Database
author:
  - name: Frank Driscoll
  - name: Cam Garfield
  - name: Cole Guerin
  - name: Hannah Hartnett
    
  
affiliation:
  - address: Statistical Graphics, Colby College
column_numbers: 4
primary_colour: '#8B0000'
output: 
  posterdown::posterdown_html:
    self_contained: false
bibliography: packages.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(Lahman)
library(dplyr)
library(reshape2)
library(ggplot2)
library(gridExtra)
library(viridis)
library(forcats)

data(People)
people <- People
head(People)

batting <- Batting
head(batting)

pitching <- Pitching
head(pitching)

fielding <- Fielding
head(fielding)

sums <- batting %>% 
  group_by(playerID) %>% 
  summarise(ab_sum = sum(AB), k_sum = sum(SO), hr_sum = sum(HR))
sums$hr_ab = sums$hr_sum/sums$ab_sum
sums$k_ab = sums$k_sum/sums$ab_sum

by_ID <- sums %>% 
  filter(ab_sum > 1000) %>%
  filter(k_ab>0) %>% 
  filter(hr_ab>0)
  
by_year <- batting %>%
  group_by(yearID) %>%
  summarise(ab = sum(AB), k = sum(SO), hr = sum(HR))

by_year$hr_ab = by_year$hr/by_year$ab
by_year$k_ab = by_year$k/by_year$ab

# establish theme of graphs
baseball_theme <-  theme_gray() + # White background, black and white theme
  theme(axis.text = element_text(size = 10, color = "azure4"),
        text = element_text(size = 10, face = "bold", color = "azure4", family = 'Palatino'))
```

# Introduction

The Major League Baseball association has served as America's pastime's highest level of competition for almost 120 years. We dive into the in-depth historical database that chronicles the sport in order to examine how hitting has changed over time.

<br>

# The Database

Lahman's Baseball Database is part of the Comprehensive R  Archive Network. The updated version of the database contains complete batting and pitching statistics from 1871 to 2020, plus fielding statistics, standings, team stats, managerial records, post-season data, and more.

<br>

# Looking at Continuous Hit Data
<div style="margin-bottom:100px;">
```{r, warning = FALSE, message  = FALSE, echo = FALSE, fig.height = 6, out.width='100%', fig.align = "center"}

# Hannah plot 1
# take continuous variables
batting_continuous <- Batting %>% select(G, AB, R, H, X2B, X3B, HR, RBI, SB, CS, BB, SO, IBB, HBP, SH, SF, GIDP)

cormat <- round(cor(batting_continuous, use = 'complete.obs'), 2)


library(reshape2)
melted_cormat <- melt(cormat)


reorder_cormat <- function(cormat){
  # Use correlation between variables as basis for distance
  dd <- as.dist(1-abs(cormat))
  hc <- hclust(dd)
  cormat <- cormat[hc$order, hc$order]
  return(cormat)
}

# Function to get lower triangle of corr matrix
get_lower_tri <- function(cormat){
  cormat[upper.tri(cormat)] <- NA
  return(cormat)
}

lo_tri_reordered <- cormat %>%
  reorder_cormat %>%
  get_lower_tri %>%
  melt


ggplot(data = lo_tri_reordered, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile() +
  labs(x = 'Batting Variable',
       y = 'Batting Variable',
       title = 'Correlation Heat Map for Continuous Batting Variables',
       fill = 'Correlation') +
  baseball_theme +
  theme(axis.text.x = element_text(angle = 45), text = element_text(size=13)) +
  scale_fill_gradient2(
  low = "#8B0000",
  mid = "white",
  high = "navy",
  midpoint = 0,
  limits = c(-1, 1))

```

We began with a high-level analysis of correlations between variables. Unsurprisingly, the number of games, hits, at bats, runs, and runs batted in are very highly positively correlated. There are no negatively correlated pairs of variables, though sacrifice hits have essentially no association with either home runs or intentional walks.

<br>
<br>
<br>



</div>

# Exploring HRs per AB

```{r, warning = FALSE, message  = FALSE, echo = FALSE, fig.height = 4, out.width='100%', fig.align = "center"}
density <- ggplot(by_year, aes(x = hr_ab))+
  geom_density(color = "navy",fill = "navy",alpha = 0.25,size = 2)+
  labs(x = "HR/AB", y = "Density of HR/AB by Year", title = "Density distribution of HR/AB from 1871-2020") + baseball_theme

density
```

The distribution of HR/AB by year has a two peaked distribution, with the first at around 0.05-- a high concentration of low HR/AB numbers and the second at around 0.025 representing a concentration of higher HR/AB values. There is a decent right skew representing the few players with higher HR/AB numbers.

<br>

# What influences number of home runs hit?

```{r, warning = FALSE, message  = FALSE, echo = FALSE, fig.height = 4, out.width='100%', fig.align = "center"}
years <- ggplot()+
  geom_line(data = by_year, aes(x = yearID, y = hr_ab, color = "blue")) +
  geom_line(data = by_year, aes(x = yearID, y = k_ab, color = "red")) +
  scale_colour_manual(name = 'Metric',
                      values =c('blue'='navy','red'='#8B0000'), labels = c('HR/AB','K/AB'))+
  labs(x = "Year", y = "Metric/AB", title = "HR/AB and K/AB by Year")
years + baseball_theme
```

Over the years, both home runs and strikeouts have increased, indicating a transition from small ball baseball
where pitchers throw to contact, to high power baseball where stronger hitters try to hit home runs and pitchers
with increased velocity go for strikeouts.

<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>

# Exploring Hit Type

```{r, warning = FALSE, message  = FALSE, echo = FALSE, fig.height = 4, out.width='100%', fig.align = "center"}
# Cole Plot 1

batting <- Batting %>%
  # select the variables that we want left after we filter the data
  select(yearID, H, X2B, X3B, HR) %>%
  # select the years from 1871+
  filter(yearID >= 1871) %>%
  group_by(yearID) %>%
  #  summarise_each(funs(sum(., na.rm=TRUE))) %>%
  summarise_all(funs(sum(., na.rm=TRUE))) %>%
  # we summarize by year, and then na.rm takes care of 0's in the data
  mutate(X1 = H-(X2B+X3B+HR)) %>% #create a column for singles
  # we eventually want these as a percentage of hits, so we can do the math now
  mutate(Single = X1/H*100) %>%
  mutate(Double = X2B/H*100) %>%
  mutate(Triple = X3B/H*100) %>%
  mutate(HomeRun = HR/H*100)

bat <- batting %>%
  select(yearID, Single, Double, Triple, HomeRun)

bat_long <- melt(bat, id.vars = c("yearID"))

hitsperyear <- ggplot(bat_long, aes(x=yearID, y= value, col=variable)) +
  geom_line() +
  xlab("Major League Baseball Season") +
  ylab("Percentage") +
  ggtitle("Hits by Type in Major League Baseball") +
  scale_x_continuous(breaks = c(1870, 1885, 1900, 1915, 1930, 1945,
                                1960, 1975, 1990, 2005, 2020 )) +
  scale_y_continuous(breaks = c(0, 25, 50, 75, 100))+
  guides(colour=guide_legend(reverse=FALSE,
                             aes(ggtitle= "Type of Hit"))) +
  baseball_theme +
  scale_colour_manual(values = c('navy', '#8B0000', 'palegreen4', 'mediumpurple3')) +
  theme(text = element_text(size=15))


hitsperyear
```

Similarly, we see an increase in homeruns and doubles hit over time, corresponding with a decrease in singles and triples. There is a particularly large increase in home runs and decrease in singles in the last ten years.


<br>

# Batting Average getting Lower?

```{r, warning = FALSE, message  = FALSE, echo = FALSE, fig.height = 6, out.width='100%', fig.align = "center"}
# Cole Plot 2
library(ggseas)

S <- summarize(group_by(Batting, yearID, playerID),
               H=sum(H), AB=sum(AB),
               BB=sum(BB), HBP=sum(HBP, na.rm=TRUE),
               SF=sum(SF, na.rm=TRUE))

ST <- summarize(group_by(Teams, yearID),
                Games = round(mean(W + L)),
                min.AB = 3 * Games)
S2 <- merge(S, ST, by="yearID")
S.regular <- filter(S2, AB >= min.AB)

SD.data <- summarize(group_by(S.regular, yearID),
                     SD.AVG = sd(H / AB),
                     SD.OBP = sd((H + BB + HBP) / (AB + BB + HBP + SF)))

sd_avg <-  ggplot(SD.data, aes(yearID, SD.AVG)) + geom_point(colour = 'navy') +
  stat_rollapplyr(width = 10, align = "right", size = 1, color = "#8B0000") +
  labs(title = "Standard Deviations of AVG of Regular Hitters",
       x = "Year", y = "SD of Batting Average") +
  baseball_theme +
  theme(text = element_text(size=15))
  

sd_obp <- ggplot(SD.data, aes(yearID, SD.OBP)) +
  geom_point(colour = 'navy') + 
  stat_rollapplyr(width = 10, align = "right", size = 1, color = "#8B0000") +
  labs(title = "Standard Deviations of OBP of Regular Hitters",
       x = "Year", y = "SD of On Base Percentage") +
  baseball_theme +
  theme(text = element_text(size=15))
grid.arrange(sd_avg, sd_obp, ncol = 1)
```


A recent study theorized that the disappearance of the 0.400 batting average in baseball is caused by the steady decrease of batting averages among “regular” players overtime, perhaps related to the earlier claim of pitchers with high velocity attempting more strikeouts. The great hitters are moving closer to the average because of this smaller variation, so fewer athletes have been able to reach the 0.400 batting average mark. In the top plot, we see that the standard deviation of batting average for so called “regular” players has decreased over time, but it has been very consistent in the last 60 years. We also see a drop in the standard deviation of on-base percentage, shown in the bottom plot.



# Hitting Statistics Rising

```{r, warning = FALSE, message  = FALSE, echo = FALSE, fig.height = 4, out.width='100%', fig.align = "center"}

# datasets for mvp and allstar awards just in case we want to use
data(AwardsPlayers)
mvp <- subset(AwardsPlayers, awardID == "Most Valuable Player")
allstar <- AllstarFull

# adding adv stats to datasets
bstats <- battingStats()
mvp <- merge(mvp, bstats, by = c("playerID", "yearID"))
allstar <- merge(allstar, bstats, by = c("playerID", "yearID"))

# mirroring the process we have performed earlier to create single, double, etc. categories for hitting dataset
Bstats <- bstats %>%
  # select the variables that we want left after we filter the data
  select(yearID, H, X2B, X3B, HR, RBI, BA, SlugPct, OBP, OPS, BABIP) %>%
  # select the years from 1871+
  filter(yearID >= 1871) %>%
  group_by(yearID) %>%
  #  summarise_each(funs(sum(., na.rm=TRUE))) %>%
  summarise_all(funs(sum(., na.rm=TRUE))) %>%
  # we summarize by year, and then na.rm takes care of 0's in the data
  mutate(X1 = H-(X2B+X3B+HR)) %>% #create a column for singles
  # we eventually want these as a percentage of hits, so we can do the math now
  mutate(Single = X1/H*100) %>%
  mutate(Double = X2B/H*100) %>%
  mutate(Triple = X3B/H*100) %>%
  mutate(HomeRun = HR/H*100)

# selecting only adv statistics from batting statistics
advStat <- Bstats %>%
  select(yearID, BABIP, OBP, SlugPct)
advStatYear <- melt(advStat, id.vars = c("yearID"))
# removed ops because that is a measure of obp and slugpct combined, also messed with the scale of the graphs

# create histograms for yearly totals of adv stats
statperyear <- ggplot(advStatYear, aes(x = yearID, y = value)) +
  geom_histogram(stat = "identity", color = "navy", alpha = 0.7) +
  geom_smooth(method = loess, se = F ,color = "red") +
  xlab("Arranged by Major League Baseball Season") +
  ylab("Cumulative Values") +
  ggtitle("Cumulative Advanced Hitting Statistics Rising Over Time in the MLB by year") +
  scale_x_continuous(breaks = c(1870, 1900,  1930,
                                1960, 1990,  2020 )) +
  guides(colour=guide_legend(reverse=TRUE,
                             aes(ggtitle= "Stat Type"))) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  facet_wrap(. ~ variable, ncol = 2)

statperyear + baseball_theme
```


<br>

```{r, warning = FALSE, message  = FALSE, echo = FALSE, fig.height = 4, out.width='100%', fig.align = "center"}

# creating datasets categorized by decade long data
mvp2 <- mutate( mvp,
                  yearID = cut(yearID, c(1890, 1900, 1910, 1920, 1930, 1940, 1950, 1960, 1970, 1980, 1990, 2000, 2010, 2020, Inf),
                             labels = c("1890s", "1900s", "1910s", "1920s", "1930s", "1940s", "1950s", "1960s", "1970s", "1980s", "1990s", "2000s", "2010s", "2020") ))

advStat2 <- mutate(advStat, 
                   yearID = cut(yearID, c(1870, 1880, 1890, 1900, 1910, 1920, 1930, 1940, 1950, 1960, 1970, 1980, 1990, 2000, 2010, Inf),
                             labels = c("1870s", "1880s","1890s", "1900s", "1910s", "1920s", "1930s", "1940s", "1950s", "1960s", "1970s", "1980s", "1990s", "2000s", "2010s") ))

advStatYear2 <- melt(advStat2, id.vars = c("yearID"))



# creating bar graph for adv stats over decade long periods (15 decades in total)
plotA <- ggplot(advStatYear2, aes(x = yearID, y = value, fill = yearID)) +
  geom_bar(stat = "identity") + 
  facet_wrap(. ~ variable, ncol = 2)

plotA + coord_polar(theta = 'x') +
  baseball_theme +
  labs(title = "Advanced Statistics Growing each Decade", x = "Sorted by specific advanced statistics", y = "Corresponding Cumulative Values", fill = "Decade") +
  theme(axis.text.x = element_text(size = 5),  axis.ticks = element_blank(), axis.text.y = element_blank()) +
  scale_fill_manual(values = c('indianred1', 'indianred2', 'indianred3', 'indianred4', 'palegreen1', 'palegreen2', 'palegreen3', 'palegreen4', 'medium purple1', 'medium purple2', 'medium purple3', 'goldenrod', 'goldenrod1', 'mediumblue', 'navy')) 
```


<br>


# Awards and Hitting Relationship

```{r, warning = FALSE, message  = FALSE, echo = FALSE, fig.height = 4, out.width='100%', fig.align = "center"}
data(AwardsPlayers, package = 'Lahman')
awards <- AwardsPlayers


awards_hits <- merge(awards, Batting, by = 'playerID')




bat <- Batting %>%
  # select the variables that we want left after we filter the data
  select(playerID, H, X2B, X3B, HR) %>%
  # select the years from 1871+
  group_by(playerID) %>%
  #  summarise_each(funs(sum(., na.rm=TRUE))) %>%
  summarise_all(funs(sum(., na.rm=TRUE))) %>%
  # we summarize by year, and then na.rm takes care of 0's in the data
  mutate(X1 = H-(X2B+X3B+HR)) #create a column for singles
  # we eventually want these as a percentage of hits, so we can do the math now


library(tidyr)
batting_long <- bat %>% select(-H) %>%
  pivot_longer(!c(playerID), names_to = "hit_type", values_to = "count")

total_hits_long <- bat %>% select(playerID, H) %>%
  pivot_longer(!c(playerID), names_to = "total_hits", values_to = "total_hit_count")

combined_long <- merge(batting_long, select(total_hits_long, -total_hits), by = 'playerID')

awards_hit_type <- merge(select(AwardsPlayers, c(awardID, playerID)), combined_long, by = 'playerID') %>% select(-playerID) %>% group_by(awardID, hit_type) %>%
  summarise_all(funs(sum(., na.rm=TRUE)))

awards_hit_type_marginal <- awards_hit_type %>%
  group_by(awardID, hit_type) %>%
  summarize(count = count,
            total = total_hit_count,
            proportion = round(count / total, 4),
            percentage = proportion * 100)


library(forcats)
awards_hit_type_marginal <- mutate(awards_hit_type_marginal,
                 hit_type = fct_recode(hit_type,
                                  "Single" = "X1",
                                  "Double" = "X2B",
                                  "Triple" = "X3B",
                                  "Home Run" = "HR"))

awards_hit_type_marginal <- awards_hit_type_marginal %>% mutate(hit_type = fct_relevel(factor(hit_type), c("Single", "Double", "Triple", "Home Run")))

ggplot(filter(awards_hit_type_marginal, awardID == 'Outstanding DH Award' | awardID == 'Cy Young Award' | awardID == 'Most Valuable Player'), aes(x = awardID, y = percentage, fill = hit_type)) +
  geom_bar(stat = 'identity') +
  labs(x = 'Award',
       y = 'Percentage of Hit Type',
       fill = 'Hit Type',
       title = 'Number of Singles Hit Varies by Award Type') +
  baseball_theme +
  scale_fill_manual(values = c('navy', '#8B0000', 'palegreen4', 'mediumpurple3'))
```

Unsurprisingly, the players who have received awards specific to hitting, like Outstanding Designated Hitter, have much higher home run totals and fewer singles than nonspecific awards, like MVP. Those players who have received awards unrelated to hitting, like the Cy Young Award for pitching, have the fewest collective home runs and the most singles. 


<br>

# Sources

Lahman's Baseball Database 

http://www.seanlahman.com/baseball-archive/statistics/


Stephen Jay Gould. Full House: The Spread Of Excellence From Plato To Darwin (New York: Harmony Books, 1996).
